---
title: "Continent-scale barcode coverage of lodgepole pine and big sagebrush"
format: html
---

This notebook will pull specimen records for two selected examples to highlight the effect of local sampling on global coverage. 

**NOTE:** Make sure GDAL is installed on the system you are running the notebooks from.
```{r install_pkgs, results='hide', message=FALSE, warning=FALSE}
## First check for the required packages, install if needed, and load the libraries.
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("sangerseqR")
remotes::install_github("ropensci/bold")
remotes::install_github("ropensci/taxize")
remotes::install_github("r-spatial/sf", configure.args = "--with-proj-lib=$(brew --prefix)/lib/")

if (!require("pacman")) install.packages("pacman")
pacman::p_load(maps, ggplot2, dplyr, countrycode, rgbif, data.table, raster, mapproj)
```

## Figure 2A-B. Get species keys for maps of *Pinus contorta* and *Artemisia tridentata* occurrences.

```{r}
# Using t.mean raster layer created above with worldclim: Change t.mean so that all cells have zero values 
t.mean.addvalues <- t.mean
class(t.mean.addvalues)
raster::values(t.mean.addvalues)[values(t.mean.addvalues) != "NA"] = 0
str(t.mean.addvalues)
plot(t.mean.addvalues) 
t.mean.addvalues
t.mean.addvalues_extent_use <- extent(t.mean.addvalues)

# Get taxon keys from GBIF
key_artemisia_tridentata <- name_suggest(q='Artemisia tridentata', rank='species')
key_artemisia_tridentata

key_pinus_contorta <- name_suggest(q='Pinus contorta', rank='species')
key_pinus_contorta
```

## Figure 2A: Map of _Pinus contorta_ GBIF records
```{r}
# Use occ_search() to query gbif records 
PINCON_occurrences_2 <- occ_search(taxonKey = 5285750, limit = 50000) # number found: 41,033
head(PINCON_occurrences_2, 2)

# Count matches: 15336
PINCON_occurrences_2_df <- data.frame(PINCON_occurrences_2$data) 

# Obtain coordinate data and project to desired CRS 
PinusSP <- data.frame(x_coords = PINCON_occurrences_2_df$decimalLongitude, y_coords = PINCON_occurrences_2_df$decimalLatitude) 
head(PinusSP, 2)
PinusSP <- subset(PinusSP, x_coords != "")
coordinates(PinusSP) <- ~x_coords + y_coords
proj4string(PinusSP) <- CRS("+proj=longlat +datum=WGS84 +no_defs")
head(PinusSP, 2)

# Check objects for plotting
head(t.mean.addvalues, 2)
PinusSP_raster <- t.mean.addvalues

# Extract cell numbers where coordinates match
matching_cells <- cellFromXY(PinusSP_raster, coordinates(PinusSP))

# Set the values in PinusSP_raster to 1 where coordinates match
PinusSP_raster[matching_cells] <- 1
head(PinusSP_raster, 2)
plot(PinusSP_raster)

PinusSP_raster_dfpoints <- rasterToPoints(PinusSP_raster)

# Convert to data frame
PinusSP_raster_df <- as.data.frame(PinusSP_raster_dfpoints)
head(PinusSP_raster_df, 2)

# Plot using ggplot 
PinusSP_raster_df <- subset(PinusSP_raster_df, layer > 0)
#PinusSP_raster_df

# ggplot wrapper for world map 
world_map <- map_data("world") #ggplot wrapper of map()
north_america_map <- subset(world_map, region == "Mexico" | region == "Canada" | region == "USA")
head(north_america_map, 2)

#Yellowstone outline
aoi_boundary_YNP <- st_read("../data/YellowstonePark/YellowstonePark1995.shp")
aoi_boundary_YNP #examine features
st_crs(aoi_boundary_YNP) #check coordinate system 
aoi_boundary_YNP_WGS84 <- st_transform(aoi_boundary_YNP,CRS("+proj=longlat +datum=WGS84")) #choose a projection and datum that every spatial object we add to the map will be converted to before plotting
st_crs(aoi_boundary_YNP_WGS84) #check coordinate system to ensure projection worked
aoi_boundary_YNP_WGS84 #re examine features

#Plot 
combtab_Pincon <- subset(combtab, species_name == "Pinus contorta")
aoi_boundary_YNP_WGS84_df <- as.data.frame(aoi_boundary_YNP_WGS84, xy = TRUE)
pinusmap2 <- ggplot() + 
  theme_classic() +
  geom_tile(data = PinusSP_raster_df, aes(x, y, fill = layer)) + 
  geom_map(data = north_america_map, map = north_america_map, aes(map_id = region), fill = "NA", color = "lightgray", size = 0.2) + 
  scale_fill_viridis_c(na.value = "white", option = "viridis", direction = 1, begin =0.6) +
  xlab("") + ylab("") + 
  scale_y_continuous(breaks = seq(-15, 75, 15), limits = c(20, 70)) +
  scale_x_continuous(breaks = seq(-160, -60, 40), limits = c(-165, -50)) +
  theme(legend.position = "none") +
  geom_point(data = combtab_Pincon, aes(x = lon, y = lat), pch = 3, size = 3, stroke = 0.9, color = alpha("black", 0.9)) + geom_sf(data = aoi_boundary_YNP_WGS84, fill = "transparent", lwd = 0.25, color = "blue")
pinusmap2
ggsave("pinusmap20240624_1.pdf", pinusmap2, width = 20, height = 8, units = "cm")
```

## Figure 2B: Map of _Artemisia tridentata_ gbif records
```{r}
# Use occ_search() to query GBIF records 
ArtTri_occurrences_2 <- occ_search(taxonKey = 9396703, limit = 20000) #16,738 found
ArtTri_occurrences_2

# Count matches
ArtTri_occurrences_df <- data.frame(ArtTri_occurrences_2$data) 

# Obtain coordinate data and project to desired CRS 
ArtemisiaSP <- data.frame(x_coords = ArtTri_occurrences_df$decimalLongitude, y_coords = ArtTri_occurrences_df$decimalLatitude) 
ArtemisiaSP
ArtemisiaSP <- subset(ArtemisiaSP, x_coords !="")
coordinates(ArtemisiaSP) <- ~x_coords + y_coords
proj4string(ArtemisiaSP) <- CRS("+proj=longlat +datum=WGS84 +no_defs")
head(ArtemisiaSP, 2)

# Check objects for plotting
head(t.mean.addvalues, 2)
ArtemisiaSP_raster <- t.mean.addvalues

# Extract cell numbers where coordinates match
matching_cells <- cellFromXY(ArtemisiaSP_raster, coordinates(ArtemisiaSP))

# Set the values in PinusSP_raster to 1 where coordinates match
ArtemisiaSP_raster[matching_cells] <- 1
head(ArtemisiaSP_raster, 2)
plot(ArtemisiaSP_raster)

ArtemisiaSP_raster_dfpoints <- rasterToPoints(ArtemisiaSP_raster)

# Convert to data frame
ArtemisiaSP_raster_df <- as.data.frame(ArtemisiaSP_raster_dfpoints)

# Plot using ggplot 
ArtemisiaSP_raster_df <- subset(ArtemisiaSP_raster_df, layer > 0)
head(ArtemisiaSP_raster_df, 2)

#Plot
combtab_Arttri <- subset(combtab, species_name == "Artemisia tridentata")
Artemisiamap3 <- ggplot() + 
  theme_classic() +
  geom_tile(data = ArtemisiaSP_raster_df, aes(x, y, fill = layer)) + 
  geom_map(data = north_america_map, map = north_america_map, aes(map_id = region), fill = "NA", color = "lightgray", size = 0.2) + 
  scale_fill_viridis_c(na.value = "white", option = "inferno", direction = -1, begin = 0.5) +
  xlab("") + ylab("") + 
   scale_y_continuous(breaks = seq(-15, 75, 15), limits = c(20, 70)) +
  scale_x_continuous(breaks = seq(-160, -60, 40), limits = c(-165, -50)) +
  theme(legend.position = "none") +
  geom_point(data = combtab_Arttri, aes(x = lon, y = lat), pch = 8, size = 3, stroke = 0.9, color = alpha("black", 0.9)) + geom_sf(data = aoi_boundary_YNP_WGS84, fill = "transparent", lwd = 0.25, color = "blue")
Artemisiamap3
ggsave("Artemisiamap4_20240626_1.pdf", Artemisiamap3, width = 20, height = 8, units = "cm")
```